\documentclass[a4paper]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{pgfplots}
\usepackage{amsmath, enumerate, amssymb, multirow, fancyhdr, color, graphicx, lastpage, listings, tikz, pdflscape, subfigure, float, polynom, hyperref, tabularx, forloop, geometry, listings, fancybox, tikz, forest, tabstackengine, cancel}
\input kvmacros
\geometry{a4paper,left=3cm, right=3cm, top=3cm, bottom=3cm}
\pagestyle{fancy}
\title{Assignment 3, Philipp Noel von Bachmann}


\begin{document}
\maketitle

\section{Problem}
    In the N-body Problem, we want to predict the movement of n bodies that
    interact with each other over time. The interaction is given by the force
    they excert on each other, which in turn changes the acceleration of each
    body, thus the velocity and position. The problem is modeled by assuming
    pairwise interaction between the bodies and taking their sum as the total
    force excerted onto each body.

\section{Solution}
    \subsection{Data structure}
        In general, we have the following 4 characteristics of each body:
        \begin{itemize}
            \item velocity
            \item position
            \item mass
            \item brightness
        \end{itemize}
        Mass and brightness are represented as double, since they are both
        continuous. As position and velocity are both vectors, they are
        represented as a struct of coordinates. Since we model the problem in
        2D, each vector thus comprises of an x and y coordinate. The
        characteristics are collected in a struct called body. This format
        also allows for easy reading of the input gal file, since we can
        directly read each block into a body struct.

        In later parts \ref{sec:mem_usage}, we also keep track of the
        accelerations of each body in an array of coordinates. We opt not to
        integrate this into the body struct to be able to retain easy writing
        and reading of files.

    \subsection{Structure}
        The baseline code consists of the following modules/functions:
        \begin{itemize}
            \item utility functions for coordinates: compute direction and compute norm 
            \item compute force: computes the force that acts onto a body
            \item compute body update: computes the update step for each body
            \item step: handles the update step, calls the subroutines. Only this function actually changes the state of our bodys.
        \end{itemize}

        In the final code, we restructured it to the following, see mostly section \ref{sec:mem_usage} for details.
        \begin{itemize}
            \item utility functions for coordinates: compute direction and compute norm 
            \item compute acceleration: computes the acceleration that acts onto a body
            \item update body: computes the update step for each body and applies it
            \item step: calls the subroutines and iterates over all bodys
        \end{itemize}


\section{Performance and Discussion}

    \subsection{Hardware information}
        \begin{itemize}
            \item CPU: Apple M1 with 4 performance cores and 4 efficiency cores
            \item GPU: Apple M1 8 core.
            \item Compiler: Slang, Version 13.0.0
        \end{itemize}

    \subsection{Performance Measurement script}
        The performance was measured with a simple shell script to enable
        reproducability. In the script we vary the number of bodies. In the end
        we also check that the code is correct with the \textbf{compare\_gals} file. Our
        time measurement was done within the code with the help of the time
        \textbf{timeoftheday} function. We only started the time measurement
        immediatly before calling the step loop and stopped the measurement
        directly after the loop.

    \subsection{Optimization and Results}
        \subsubsection{Baseline}
            \noindent\begin{tabular}{c|c|c|c}
                steps & 10 & 100 & 1000 \\
                \hline
                time (s) & 0.007 & 0.065 & 0.648\\
            \end{tabular}

            \noindent\begin{tabular}{c|c|c|c|c}
                bodies & 10 & 100 & 500 & 1000 \\
                \hline
                time (s) & 0.003 & 0.065 & 1.615 & 6.502\\
            \end{tabular}
            
            Here, we see that our computation time scales linearly with the
            number of steps, which we would expect since steps are in an outer
            loop. In comparison, we see that the number of bodies doesn't scale
            linearly with time. This makes sense since our naive algorithm has complexity $O(n^2)$.

        \subsection{Using optimizer flags}
        \begin{itemize}
            \item -O2
                \noindent\begin{tabular}{c|c|c|c|c}
                    bodies & 10 & 100 & 500 & 1000 \\
                    \hline
                    time (s) & 0.000 & 0.012 & 0.291 & 1.164\\
                \end{tabular}
            \item -O3
                \noindent\begin{tabular}{c|c|c|c|c}\label{tab:baseO3}
                    bodies & 10 & 100 & 500 & 1000 \\
                    \hline
                    time (s) & 0.001 & 0.012 & 0.290 & 1.163\\
                \end{tabular}
            \item -03 -ffast-math
                \noindent\begin{tabular}{c|c|c|c|c}
                    bodies & 10 & 100 & 500 & 1000 \\
                    \hline
                    time (s) & 0.000 & 0.002 & 0.037 & 0.146\\
                \end{tabular}
        \end{itemize}

        We can see that by enabling optimization flags, we can improve the
        performace of our code significantly. If we test our problem on
        ellipse\_N\_00100\_after200steps.gal, even enabling ffast-math does not
        change the correctness of the result. Since we might loose precision in
        general however, we opt to only use -O3 in all further
        optimizations.


        \subsection{Optimizing Instructions}
            We can make the following small optimizations:
            \begin{itemize}
                \item Remove division by $m_i$ for $a_i$ by also removing it in $F_i$, however this doesn't change the computation time.
                \item Give inline hint to our vector utility function like norm, however no improvement since compiler does it probably by itself.
            \end{itemize}

            We can reduce the time of computing $r^3$ by explicitly writing it
            out instead of using the \textbf{pow} function. This leads to a big improvement:
            \newline
            \noindent\begin{tabular}{c|c|c|c|c}
                bodies & 10 & 100 & 500 & 1000 \\
                \hline
                time (s) & 0.000 & 0.002 & 0.038 & 0.151\\
            \end{tabular}


        \subsection{Memory optimization}
            \subsubsection{Reducing memory usage}\label{sec:mem_usage}
                Instead of creating a new array of bodies for each timestep, we
                can also update our bodies in place. This however means that we
                need to restructure our code such that we first compute all
                accelerations, and then updating them. Surprisingly this doesn't
                lead to a big performance improvement, our time reduced to 0.149
                for 1000 bodys which is just 2 ms. One of the reasons could be
                that while we create fever objects, we need to split our update
                loop in two on order to first compute all accelerations and then
                compute all body updates.

            \subsubsection{Optimizing cache usage}\label{sec:cache_blocking}
                We can see in the formula for force
                \begin{equation}\label{for:force}
                    F_i = -Gm_i \sum_{j\neq i} \frac{m_j}{r^2_{ij}}\hat{r}_ij
                \end{equation}
                that we our sum goes over all $j$. If we have a very big number
                of bodies, the first body in the cache will be replaced by the
                time we come to the end of the sum. Thus if we want to acces the
                first body in the next iteration again, we need to re-read this
                line. Instead, we can also compute the sum block-wise over all
                $j$s. However this reformatting doesn't lead to an improvement,
                instead it slows down for smaller block sizes. One reason could
                be that by adding blocks we added more for loops which make it
                harder for the compiler to optimize the code.

        \subsection{Instruction Level Parallelism}
            We can get rid of the if $i \neq j$  statement in the $F_i$ \ref{for:force}
            computation loop by dividing the loop up. This leads to a minor
            improvement of 0.006s for 1000 bodies.
        
        \subsection{Final code time measurement}
            In the end, we ended up with a code that included all optimizations
            except of \ref{sec:cache_blocking}. This lead to the following
            performance:

            \begin{tikzpicture}
 
                \begin{axis}[
                    xmin = 10, xmax = 10000,
                    ymin = 0, ymax = 15.0,
                    xtick distance = 1000,
                    ytick distance = 1,
                    grid = both,
                    minor tick num = 1,
                    major grid style = {lightgray},
                    minor grid style = {lightgray!25},
                    width = 0.8\textwidth,
                    height = 0.8\textwidth,
                    xlabel = {Number of bodies},
                    ylabel = {Time (s)},]
                 
                 
                % Plot data from a file
                \addplot[
                    scatter,
                    thin,
                    red,
                    dashed
                ] file[skip first] {results.dat};
                 
                \end{axis}
                 
            \end{tikzpicture}

            We can clearly see that this curve resembles an polynomial
            increasing cost, $O(n^2)$.
        


\section{References}
    \begin{itemize}
        \item For the timing, the \textbf{get\_wall\_seconds} from Lab06, exercise 7
        \item For data reading, the function \textbf{read\_doubles\_from\_file} from \textbf{compare\_gal\_files}, but slightly modified
    \end{itemize}



\end{document}

